### 2023-01-31 Strydom_et_al_2023_graph_embedding_preprint.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@771317a3716ac0d55a20b4d8398e9866760ad3f4 🚀
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@fb8d110628b0a1dd021f89530054474415ba9ba0 🚀

In large parts of the world, boundaries that delineate geographic
regions are merely a reflection the legacy of settler colonialism, which
drives global disparity in capacity to collect and publish ecological
data. Applying any embedding to biased data does not debias them, but
rather embeds these biases, propagating them to the models using
embeddings to make predictions. Furthermore, the use of ecological data
itself is not an apolitical act (Nost \& Goldstein, 2021): data
infrastructures tend to be designed to answer questions within national
boundaries (therefore placing contingencies on what is available to be
embedded), their use often drawing upon, and reinforcing, territorial
statecraft (see \emph{e.g.} Barrett, 2005). As per Machen \& Nost
(2021), these biases are particularly important to consider when
knowledge generated algorithmically is used to supplement or replace
human decision-making, especially for governance (\emph{e.g.} enacting
conservation decisions on the basis of model prediction). As information
on networks is increasingly leveraged for conservation actions (see
\emph{e.g.} Eero et al., 2021; Naman et al., 2022; Stier et al., 2017),
the need to appraise and correct biases that are unwittingly propagated
to algorithms when embedded from the original data is immense. These
considerations are even more urgent in the specific context of
biodiversity data. Long-term colonial legacies still shape taxonomic
composition to this day (Lenzner et al., 2022; Raja, 2022), and much
shorter-term changes in taxonomic and genetic richness of wildlife
emerged through environmental racism (Schmidt \& Garroway, 2022). These
considerations mean that ``the set of species found at a place'' is not
only a response to ecological processes, and therefore there may exist
forcing on the interactions between these species.

@@

Predictive approaches in ecology, regardless of the scale at which they
are deployed and the intent of their deployment, originate in the
framework that contributed to the ongoing biodiversity crisis (Adam,
2014) and reinforced environmental injustice (Choudry, 2013; Domínguez
\& Luoma, 2020). The risk of embedding this legacy in our models is
real, especially when the impact of this legacy on species pools is
being increasingly documented. This problem can be addressed by
re-framing the way we interact with models, especially when models are
intended to support conservation actions. Particularly on territories
that were traditionally stewarded by Indigenous people, we must
interrogate how predictive approaches and the biases that underpin them
can be put to task in accompanying Indigenous principles of land
management (Eichhorn et al., 2019; No'kmaq et al., 2021). The discussion
of ``algorithm-in-the-loop'' approaches that is now pervasive in the
machine learning community provides examples of why this is important.
Human-algorithm interactions are notoriously difficult and can yield
adverse effect (Green \& Chen, 2019; Stevenson \& Doleac, 2021),
suggesting the need to systematically study them for the specific
purpose of, here, biodiversity governance. Improving the algorithmic
literacy of decision makers is part of the solution (\emph{e.g.} Lamba
et al., 2019; Mosebo Fernandes et al., 2020), as we can reasonably
expect that model outputs will be increasingly used to drive policy
decisions (Weiskopf et al., 2022). Our discussion of these approaches
need to go beyond the technical and statistical, and into the governance
consequences they can have. To embed data also embeds historical and
contemporary biases that acted on these data, both because they shaped
the ecological processes generating them, and the global processes
leading to their measurement and publication. For a domain as vast as
species interaction networks, these biases exist at multiple scales
along the way, and a challenge for prediction is not only to develop (or
adopt) new quantitative tools, but to assess the behavior of these tools
in the proper context.

>>>>

In large parts of the world, boundaries that delineate geographic
regions are merely a reflection the legacy of settler colonialism, which
drives global disparity in capacity to collect and publish ecological
data. Applying any embedding to biased data does not debias them, but
rather embeds these biases, propagating them to the models using
embeddings to make predictions. Furthermore, the use of ecological data
itself is not an apolitical act (Nost \& Goldstein, 2021): data
infrastructures tend to be designed to answer questions within national
boundaries (therefore placing contingencies on what is available to be
embedded), their use often drawing upon, and reinforcing, territorial
statecraft (see \emph{e.g.} Barrett, 2005). As per Machen \& Nost
(2021), these biases are particularly important to consider when
knowledge generated algorithmically is used to supplement or replace
human decision-making, especially for governance (\emph{e.g.} enacting
conservation decisions on the basis of model prediction). As information
on networks is increasingly leveraged for conservation actions (see
\emph{e.g.} Eero et al., 2021; Naman et al., 2022; Stier et al., 2017),
the need to appraise and correct biases that are unwittingly propagated
to algorithms when embedded from the original data is immense. These
considerations are even more urgent in the specific context of
biodiversity data. Long-term colonial legacies still shape taxonomic
composition to this day (Lenzner et al., 2022; Raja, 2022), and much
shorter-term changes in taxonomic and genetic richness of wildlife
emerged through environmental racism (Schmidt \& Garroway, 2022). Thus,
the set of species found at a specific location is not only as the
result of a response to ecological processes separate from human
influence, but also the result of human-environment interaction as well
as the result legislative/political histories.

@@

Predictive approaches in ecology, regardless of the scale at which they
are deployed and the intent of their deployment, originate in the
framework that contributed to the ongoing biodiversity crisis (Adam,
2014) and reinforced environmental injustice (Choudry, 2013; Domínguez
\& Luoma, 2020). The risk of embedding this legacy in our models is
real, especially when the impact of this legacy on species pools is
being increasingly documented. This problem can be addressed by
re-framing the way we interact with models, especially when models are
intended to support conservation actions. Particularly on territories
that were traditionally stewarded by Indigenous people, we must
interrogate how predictive approaches and the biases that underpin them
can be put to task in accompanying Indigenous principles of land
management (Eichhorn et al., 2019; No'kmaq et al., 2021). The discussion
of ``algorithm-in-the-loop'' approaches that is now pervasive in the
machine learning community provides examples of why this is important.
Human-algorithm interactions are notoriously difficult and can yield
adverse effects (Green \& Chen, 2019; Stevenson \& Doleac, 2021),
suggesting the need to systematically study them for the specific
purpose of, here, biodiversity governance. Improving the algorithmic
literacy of decision makers is part of the solution (\emph{e.g.} Lamba
et al., 2019; Mosebo Fernandes et al., 2020), as we can reasonably
expect that model outputs will be increasingly used to drive policy
decisions (Weiskopf et al., 2022). Our discussion of these approaches
need to go beyond the technical and statistical, and into the governance
consequences they can have. To embed data also embeds historical and
contemporary biases that acted on these data, both because they shaped
the ecological processes generating them, and the global processes
leading to their measurement and publication. For a domain as vast as
species interaction networks, these biases exist at multiple scales
along the way, and a challenge for prediction is not only to develop (or
adopt) new quantitative tools, but to assess the behavior of these tools
in the proper context.
